# Intro to ggplot {#ggplot-intro}

## Goals for this section

- An introduction to the Grammar of Graphics
- We'll make charts!

## Introduction ggplot

[ggplot2](https://ggplot2.tidyverse.org/) is the data visualization library within Hadley Wickham's [tidyverse](https://www.tidyverse.org/). It uses a concept called the [Grammar of Graphics](https://byrneslab.net/classes/biol607/readings/wickham_layered-grammar.pdf), the idea that you can build every graph from the same components: a data set, a coordinate system, and geoms -- the visual marks that represent data points.

### What I like/dislike about ggplot

The ggplot system allows you to display data right in your notebook. It is really good at helping you find important things in your data that can inform reporting. It's an important tool in your R-based data journalism toolkit.

What ggplot is less good at is creating publishable graphics. Don't get me wrong ... you can do it, but nuances of the ggplot system take time to master at that level. There are other tools (like [Datawrapper](https://www.datawrapper.de/) and [Flourish](https://flourish.studio/)) that can do that better, even without code. That said, there is a place for R in that workflow, too. We'll cover using Datawrapper in a later chapter.

### The Grammar of Graphics

ggplot uses this concept of the Grammar of Graphics ... i.e., that you can use code to describe how to build a chart layer-by-layer.

With a hat tip to [Matt Waite](http://www.mattwaite.com/), we can describe the components of the Grammar of Graphics as:

- **data**: which data you are pulling from for the chart.
- **aesthetics**: describes how to apply specific data to the plot. What is on x axis, what is on y axis, for starters.
- **geometries**: the shape the data is going to take. lines, columns, points.
- **scales**: any transformations we might make on the data.
- **layers**: how we might layer multiple geometries over top of each other to reveal new information.
- **facets**: how we might graph many elements of the same data set in the same space.

What to remember here is this: for every graphic we start with the data, and then we build a chart from it one "layer" at a time.

The best way to learn this system is to do it and explain along the way.

## Start a new project

1. Get into RStudio and make sure you don't have any other files or projects open.
1. Create a new project, name it `yourname-ggplot` and save it in your rwd folder.
1. No need to create our folder structure ... we won't need it here.
1. Start a new RMarkdown notebook and save it as `01-intro-ggplot.Rmd`.
1. Remove the boilerplate and create a setup section that loads `library(tidyverse)`, like we do with every notebook.

```{r setup, echo=F, message=F, warning=F}
library(tidyverse)
```

## The layers of ggplot

> Much of this first plot explanation comes from Hadley Wickham's [R for Data Science](https://r4ds.had.co.nz/data-visualisation.html?q=aes#data-visualisation), with edits to fit the lesson here.

We're going to use a data set that is part of the tidyverse to explore how ggplot works.

1. Start a new section "First plot" and add a code chunk.
1. Add the code below and run it to see what the mpg dataset looks like.

```{r}
mpg
```

The `mpg` data contains observations collected by the US Environmental Protection Agency on 38 models of cars. It's a data set embedded into the tidyverse for lessons like this one.

Among the variables in mpg are:

- `displ`, a car’s engine size, in liters.
- `hwy`, a car’s fuel efficiency on the highway, in miles per gallon (mpg).

With these two variables we can test the theory that cars with smaller engines (`displ`) get better gas mileage (`hwy`).

### Build the base layer

With ggplot2, you begin a plot with the function` ggplot()`. `ggplot()` creates a coordinate system that you can add layers to. The first argument for `ggplot()` is the data set to use in the graph So `ggplot(mpg)` creates an empty graph with no axes or anything.

You complete your graph by adding one or more layers to `ggplot()`. The function `geom_point()` adds a layer of points to your plot, which creates a scatterplot. ggplot2 comes with many geom functions that each add a different type of layer to a plot.

Each geom function in ggplot2 takes a `mapping` argument. This defines how variables in your dataset are mapped to visual properties. The `mapping` argument is always paired with `aes()`, and the `x` and `y` arguments of `aes()` specify which variables to map to the x and y axes. ggplot2 looks for the mapped variables in the `data` argument, in this case, `mpg.`

We can apply `aes()` mappings to the graph as a whole and/or to the individual geom layers.

Frankly, it is easier to show this than explain it. The code below sets up the grid and axes lines, but it hasn't placed any data on the plot.

Do this:

1. Add some text that you are building the mpg chart.
2. Add the code chunk below and run it.

```{r}
ggplot(mpg, aes(x = displ, y = hwy))
```

Let's work through the code above:

- `ggplot` is our function to make a chart.
- The first argument `ggplot` needs is the data. It could be specified as `data = mpg` but we don't need the `data = ` part as it is always the first item specified inside of (or piped into) `ggplot()`
- Next is the **aesthetics** or `aes()`. This is where tell ggplot what data us plot on the `x` and `y` axis. You might see this as `mapping = aes(<VALUES>)` but we can often get by without the `mapping =` part.

In our case we are applying these `aes()` to the entire chart. You'll see later we can also specify different `aes()` to specific geoms.

### Layers can we add to our plots

We'll now add onto this base layer a number of things:

- **geometries** (or geoms as we call them) are the way we plot data on the base grid. There are [many geoms](https://github.com/rstudio/cheatsheets/blob/master/data-visualization.pdf), but here are a few common ones:
    - `geom_points()` adds dots onto the grid based on the data. Will will use these here to build at scatterplot graph.
    - `geom_line()` adds lines between data points on the grid. Basically a line chart.
    - `geom_col()` and `goem_bars()` adds bars to the grid based on values in the data. A bar chart. We'll use `geom_col()` later in this lesson but you can read about the difference between the two in a later chapter.
    - `geom_text()` adds labels based on values in the data.
- **labels** (or labs, since we use the `labs()` function for them) are a series of text-based items we can layer onto our plots like titles, bylines and axis names.
- **themes** change the visual styles of the grids and axis. There are several available within ggplot and [many other from the R community](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

We add layers onto the chart using the `+` at the end of a line. Think of the `+` as the ` %>% ` of ggplot.

### Add geom_points

1. EDIT your plot chunk to add the `+` and a new line for `geom_point()`

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + # don't forget the + at the end of this line
  geom_point() # the geom_point
```

The `geom_point()` function above is inheriting the `aes()` values from the line above it.

### Adding other mappings

We can add aesthetics to either the plot as a whole (which we did with the x and y values above) and those will apply to all the geoms unless overwritten.

But we can also add aesthetics to specific geoms. We'll demonstrate this below.

1. Edit your `geom_point()` function to add a color mapping to the points with `aes(color = class)`. `color` is the type of aesthetic, and `class` is a column in the data.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) # this is the line you are editing

```

As you can see, the dots were given colors based on the values in the `class` column, and ggplot also added a legend to the graphic.

There are other aesthetics you can use.

1. Change the `color` aesthetic to one of these values and run it to see how it affects the chart: `alpha`, `size` and `shape`. (i.e., `alpha = class`.)
2. Once you've tried them, change it back to `color`.

OK, enough of the basics ... let's build a chart you _might_ care about.

## Let's build a bar chart

We'll build some charts from our first-day survey where you told me your favorite Disney Princess and favored flavor of ice cream.

We aren't going to create different notebooks or download the data to to your computer ... we're just doing to save it directly into a tibble.

1. Start a new section: Princess preference chart.
1. Use text to note that we'll use class data to build a chart.
1. Add the code below to get the data.

```{r class-data}
# read the data and fill the tibble: class
class <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRnSAx9eBoOGdZ3pMLZ2XhrBlgl56BeycxJwdTofmgfTBTZ7A1-LMuBxAI094aAZnCmeThPNXaU-xro/pub?gid=1648328850&single=true&output=csv")

# peek at the data
class
```

### Prepare the data

While there are ways for ggplot to calculate values from your data on the fly, I much prefer to first build a table of the values I want plotted on a chart.

Our goal here is to make a bar (or column) chart showing the number of votes for each princess from the data. So, we need to count the number of rows for each value ... our typical group_by/summarize/arrange process. I'm going to use the `count()` shortcut for our GSA here since we haven't used it much lately. I'm saving the summarized data into a new dataframe called `princess_data`. Follow along in your notebook:

1. Add a new section: Princess chart
1. Add text that you are creating a data frame to plot.
2. Add the code below to create that data.

```{r princess-summarize}
princess_data <- class %>% 
  count(princess, name = "votes", sort = TRUE)
  # this above line counts the princess rows, sets the name and sorts

# peek at the data
princess_data
```

I hope you understand what we've done here. We're counting the number of rows for each princess.

### Build our plot with geom_col

Our first goal is to build the first layer the plot ... basically tell ggplot what data we are using so it will build the grid to hold our plots.

1. Add some text noting that you'll now plot.
1. Add the following code chunk that starts the plot and run it.

```{r}
princess_data %>% 
  ggplot(aes(x = princess, y = votes)) # sets our x and y axes
```

Note that I used ` %>% ` to pipe the data into `ggplot()` instead of naming it as an argument.

You'll see the grid and x/y axis of the data, but no geometries are applied yet. Don't worry yet now it looks ... we'll get there.

### Add the geom_col layer

Now it is time to add our columns.

1. Edit the plot code to add the ggplot pipe `+` and on the next line add `geom_col()`.

```{r}
princess_data %>% 
  ggplot(aes(x = princess, y = votes)) + # don't forget the +
  geom_col() # adds the bars
```

This added our data to the plot, though there are a couple of issues:

- We can't read the value names. We can fix this.
- The order of the bars is alphabetical instead of in vote order. Again, we can fix it.

### Flip the axis

We can "flip" the axis to turn it sideways to read the labels. This can be a bit confusing later because the "x" axis is now going up/down.

1. Edit your plot chunk to add the ggplot pipe `+`and `coord_flip()` on the next line.

```{r princess-flipped}
princess_data %>% 
  ggplot(aes(x = princess, y = votes)) +
  geom_col() + # don't forget the +
  coord_flip() # flips the axis
```

### Reorder the bars

The bars on our chart are in alphabetical order of the x axis (and reversed thanks to our flip.) We want to order the values based on the `votes` in the data.

We can do that by editing the `x` values in our `aes()` using the a function called `reorder()`.

The `reorder()` takes two arguments: The column to reorder, and the column to base that reorder on. It can happen in two different ways, and I'll be honest and say I don't know which is easier to comprehend.

- `x = reorder(princess, votes)` says "set the x axis as `princess`, but order as `votes`. OR ...
- `x = princess %>% reorder(votes)` says "set the x axis as `princess` _and then_ reorder by `votes`.

They both work. Even though I'm a fan of the tidyverse ` %>% ` construct, I'm going with the first version.

```{r}
princess_data %>% 
  ggplot(aes(x = reorder(princess, votes), y = votes)) + # this is the line you edit
  geom_col() +
  coord_flip()
```

### Add some titles, labels

Now we'll add a **layer** of labels to our chart using the the `labs()` function. You'll see we can add and change a number of things with `labs()`.

```{r princess-labs}
princess_data %>% 
  ggplot(aes(x = reorder(princess, votes), y = votes)) + 
  geom_col() +
  coord_flip() + # don't forget +
  labs(
    title = "Favored princess", # adds a title
    subtitle = "Disney Princess votes from Reporting with Data, Fall 2021.", # adds a subtitle
    caption = "By Christian McDonald", # adds the byline
    x = "Princess choices", # renames the x axis label (which is really y since it is flipped)
    y = "Number of votes" # renames the y axis label (which is really x since it is flipped)
  )
```

There you go! You've made a chart showing how our classes rated Disney Princesses.

## Build the ice cream chart on your own

Now it is time for you to put these skills to work:

1. Build a chart about the favorite ice creams from RWD classes.

Some things to consider:

- You need a new section, etc.
- Your starting with the same `class` data
- You need to prepare the data
- You need to build the chart

It's essentially the same process we used for the princess chart, but using `ice_cream` variable.

## Turn in this notebook

Make sure this notebook runs all the way through (Restart R and Run All), knit, stuff, and turn into Canvas.


## What we've learned

There is a ton, really.

- ggplot2 (which is really the `ggplot()` function) is the charting library for the tidyverse. This whole lesson was about it.

Here are some more references for ggplot:

- [The ggplot2 documentation](http://ggplot2.tidyverse.org/reference/index.html) and [ggplot2 cheatsheets](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf).
- [R for Data Science, Chap 3.](https://r4ds.had.co.nz/data-visualisation.html) Hadley Wickam dives right into plots in his book.
- [R Graphics Cookbook](https://r-graphics.org/) has lots of example plots. Good to harvest code and see how to do things.
- [The R Graph Gallery](https://www.r-graph-gallery.com/) another place to see examples.
