# Deeper into ggplot {#ggplot-more}

In the last chapter you were introduced to [ggplot2](https://ggplot2.tidyverse.org/index.html), the graphic function that is part of the tidyverse. With this chapter we'll walk through more ways to use ggplot. This will also serve as a reference for you. A huge hat tip to [Jo Lukito](https://www.jlukito.com/). One of her lessons inspired most of this.

## References

ggplot2 has a LOT to it and we'll cover only the basics. Here are some references you might use:

- [ggplot cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf)
- [R for Data Science](https://r4ds.had.co.nz/index.html)
- [R Graphics Cookbook](https://r-graphics.org/)
- [The R Graph Gallery](https://www.r-graph-gallery.com/) another place to see examples.
- [ggplot2: Elegant graphics for Data Analysis](https://ggplot2-book.org/index.html)


## Learning goals for this chapter

Some things we'll touch on concerning ggplot:

- Layering multiple geoms
- Adding/changing aesthetics in layers
- Using themes to change the look of our charts
- Facets (multiple charts from same data)
- Saving files
- Interactivity with Plotly

## Set up your notebook

We'll use the same `yourname-ggplot` project we used in the last chapter, but start a new RNotebook.

1. Open your plot project.
2. Start a new RNotebook. Add the goals listed above.
3. Load the tidyverse package.

```{r setup, echo=F, message=F, warning=F}
library(tidyverse)
```

### Let's get the data

> I hope to demonstrate in class the creation of this first plot. Otherwise you should be able to follow along in the screencast.

Again, we won't download the data ... we'll just import it and save it to a tibble. We are using data from a weekly project called #tidytuesday that the community uses to practice R. Perhaps in the near future we'll have our own #tidytuesday sessions!

1. Start a new section to indication you are importing the data
2. Note in text it is from #tidytuesday
3. Add the code chunk below, which will download a saved copy of the data.

```{r import}
kids_data <- read_rds("https://github.com/utdata/rwdir/blob/main/data-raw/kids-data.rds?raw=true")

# peek at the table
kids_data

# glimpse it
kids_data %>% glimpse()
```

If you want to learn more about this dataset you can find [information here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-15/readme.md). In short: "This dataset provides a comprehensive accounting of public spending on children from 1997 through 2016." Included is spending for higher education (the `highered` values in the column `variable`.)

We'll filter this data to get to our data of interest: How much does Texas (and some neighboring states) spend on higher education.

## Make a line chart of the Texas data

Our first goal here is to plot the "inflation adjusted spending per child" for higher education in Texas.

### Prepare the data

We need to filter our data to include just the `highered` values for Texas. We're going to save that filtered data into a new tibble.

1. Add a new section indicating we are building a chart to show higher education spending in Texas.
2. Note that we are preparing the data.
3. Add the chunk below and run it.

```{r tx-hied-data}
tx_hied <- kids_data %>%
  filter(
    variable == "highered",
    state == "Texas"
  )

# peek at the data
tx_hied
```

### Plot the chart

I want you to create the plot here one step at a time so you can review how the layers are added.

1. Add and run the ggplot() line first (but without the `+`)
2. Then add the `+` and the `geom_point()` and run it.
3. Then add the `+` and `geom_line()` and run it.
4. When you add the `labs()` add all the lines at once and run it.

```{r tx-hied-chart}
ggplot(tx_hied, aes(x = year, y = inf_adj_perchild)) + # we create our graph
  geom_point() + # adding the points
  geom_line() +  # adding the lines between points
  labs(
    title = "School spending slips", 
    subtitle = "Texas spent less per child on higher education in 2016.",
    x = "Year", y = "$ per child (Adjusted for Inflation)",
    caption = "Source: tidykids"
  )
  # labs above add text layer on top of graph
```

We have a pretty decent chart showing `year` on our x axis and `inf_adj_perchild` (or inflation-adjusted spending per child) on our y axis.

## Adding more information

OK, our Texas higher education spending is fine ... but how does that compare to neighboring states? Let's work through building a new chart that shows all those steps.

### Prepare the data

We need to go back to our original `kids_data` to get the additional states.

1. Start a new section that notes we are building a chart for five states.
2. Note that we'll first prepare the data.

```{r five-hied-data}
five_hied <- kids_data %>% 
  filter(
    variable == "highered",
    state %in% c("Texas", "Oklahoma", "Arkansas", "New Mexico", "Louisiana")
  )

five_hied
```

Note we used our `%in%` filter to get any state listed in `c()`.

### Plot multiple line chart

Let's add a different line for each state. To do this you would use the color aesthetic `aes()` in the `geom_line()` geom. Recall that geoms can have their own `aes()` variable information. This is especially useful for working with a third variable (like when making a stacked bar chart or line plot with multiple lines). Notice that the color aesthetic (meaning that it is in aes) takes a variable, not a color. You can learn how to change these colors [here](http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually).

1. Add a note that we'll now build the chart.
2. Add the code chunk below and run it. Look through the comments so you understand it.

```{r five-hied-lcolor}
ggplot(five_hied, aes(x = year, y = inf_adj_perchild)) + 
  geom_point() +
  geom_line(aes(color = state)) + # The aes selects a color for each state
  labs(
    title = "Spending on Higher Education in Texas, Bordering States",
    x = "Year",
    y = "$ per child (Adjusted for Inflation)",
    caption = "Source: tidykids"
  )
```

Notice that R changes the color of the line, but not the point? This is because we only included the aesthetic in the `geom_line()` geom and not the `geom_point()` geom.

1. Edit your `geom_point()` to add `aes(color = state)`.

```{r five-hied-pcolor}
ggplot(five_hied, aes(x = year, y = inf_adj_perchild)) +
  geom_point(aes(color = state)) + # add the aes here
  geom_line(aes(color = state)) +
  labs(title = "Spending on Higher Education in Texas, Bordering States", 
       x = "Year", y = "$ per child (Adjusted for Inflation)",
       caption = "Source: tidykids")
```

## On your own: Line chart

I want you to make a line chart of preschool-to-high-school spending (the "PK12ed" value in the `variable` column) showing the inflation adjusted per-child spending (the `inf_adj_perchild` column) for the six states that border the Gulf of Mexico. This is very similar to the chart you just made, but with different values.

Some things to do/consider:

1. Do this in a new section and explain it.
1. You'll need to prepare the data just like we did above to get the right data points and the right states.
2. I really suggest you build both chunks (the data prep and the chart) one line at a time so you can see what each step adds.
1. Save the resulting plot into a new R object because we'll use it later.

```{r homework-chart-example, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
ed_gulf <- kids_data %>% 
  filter(variable == "PK12ed",
         state %in% c("Texas", "Louisiana", "Mississippi", "Alabama", "Georgia", "Florida"))

ed_gulf_plot <- ed_gulf %>% 
  ggplot(aes(x = year, y = inf_adj_perchild)) +
  geom_line(aes(color = state)) +
  labs(
    title = "A good headline",
    subtitle = "A good subtitle. Also fix your axis names!!!",
    caption = "Your name"
  )
```

## Tour of some other adjustments

You don't have to add these examples below to your own notebook, but here are some examples of other things you can control.

### Line width

```{r line-width}
ggplot(five_hied, aes(x = year, y = inf_adj_perchild)) +
  geom_point(aes(color = state)) +
  geom_line(aes(color = state), size = 1.5) + # added size here
  labs(title = "Spending on Higher Education in Texas, Bordering States", 
       x = "Year", y = "$ per child (Adjusted for Inflation)",
       caption = "Source: tidykids")
```

### Line type

This example removes the points and adds a `linetype = state` to the ggplot aesthetic. This gives each state a different type of line. We also set the color in the `geom_line()`

```{r line-type}
ggplot(five_hied, aes(x = year, y = inf_adj_perchild, linetype = state)) +
  geom_line(aes(color = state), size = .75) +
  labs(title = "Spending on Higher Education in Texas, Bordering States", 
       x = "Year", y = "$ per child (Adjusted for Inflation)",
       caption = "Source: tidykids")
```

### Adjust axis

`ggplot()` typically makes assumptions about scale. Sometimes, you may want to change it though (e.g., make them a little larger). There are a couple different ways to do this. The most straightforward may be `xlim()` and `ylim()`.

```{r adjust-axis}
ggplot(five_hied, aes(x = year, y = inf_adj_perchild, linetype = state)) +
  geom_line(aes(color = state), size = .75) +
  xlim(1995, 2020) + # sets minimum and maximum values on axis
  labs(title = "Spending on Higher Education in Texas, Bordering States", 
       x = "Year", y = "$ per child (Adjusted for Inflation)",
       caption = "Source: tidykids")

```

The function `xlim()` and `ylim()` are shortcuts for `scale_x_continuous()` and `scale_y_continuous()` which [do more things](https://ggplot2.tidyverse.org/reference/scale_continuous.html#examples).

## Saving plots as an object

Sometimes it is helpful to push the results of a plot into an R object to "save" those configurations. You can continue to add layers after, but don't have to rebuild the basic chart each time. We'll do that here so we can explore themes next.

1. Edit your Texas plot chunk you made earlier to save it into an R object, and then call `tx_plot` after it so you can see it.

```{r tx-plot-create}
# the line below pushes the graph results into tx_plot
tx_plot <- ggplot(tx_hied, aes(x = year, y = inf_adj_perchild)) +
  geom_point() +
  geom_line() +
  labs(
    title = "School spending slips", 
    subtitle = "Texas spent less per child on higher education in 2016.",
    x = "Year", y = "$ per child (Adjusted for Inflation)",
    caption = "Source: tidykids"
  )

# Since we saved the plot into an R object above, we have to call it again to see it.
# We save graphs like this so we can reuse them.
tx_plot
```

We can continue to build upon the `tx_plot` object like we do below with themes, but those changes won't be "saved" into the R environment unless you assign it to an R object.

## Themes

The _look_ of the graph is controlled by the theme. There are a number of preset themes you can use. Let's look at a couple.

1. Create a new section saying we'll explore themes
2. Add the chunk below and run it.

```{r theme-minimal}
tx_plot +
  theme_minimal()
```

This takes our existing `tx_plot` and then applies the `theme_minimal()` look to it.

There are a number of themes built into ggplot, most are pretty simplistic.

1. Edit your existing chunk to try different themes. Some you might try are `theme_classic()`, `theme_dark()` and `theme_void()`. 

### More with ggthemes

There are a number of other packages that build upon `ggplot2`, including [`ggthemes`](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

1. In your R console, install the ggthemes package: `install.packages("ggthemes")`
2. Add the `library(ggthemes)` at the top of your current chunk.
3. Update the theme line to view some of the others options noted below.

```{r theme-economist}
library(ggthemes)
tx_plot +
  theme_economist()
```

```{r theme-538}
tx_plot +
  theme_fivethirtyeight()
```

```{r theme-stata}
tx_plot +
  theme_stata()
```

### There is more to themes

There is also a `theme()` function that allows you individually [adjust about every visual element](https://ggplot2.tidyverse.org/reference/theme.html) on your plot.

We may introduce more about those later in the course.

## Facets

Facets are a way to make multiple graphs based on a variable in the data. There are two types, the `facet_wrap()` and the `facet_grid()`. There is a good explanation of these in [R for Data Science](https://r4ds.had.co.nz/data-visualisation.html?q=facet#facets).

We'll start by creating a base graph and then apply the facet. 

1. Start a new section about facets
2. Add the code below to create your chart and view it.

```{r facet-data}
five_plot <- ggplot(five_hied, aes(x = year,
                                   y = inf_adj_perchild)) +
  geom_line(aes(color = state)) +
  geom_point(aes(color = state)) +
  labs(title = "Spending on Higher Education in Texas, Bordering States", 
       x = "Year", y = "$ per child (Adjusted for Inflation)",
       caption = "Source: tidykids")

five_plot
```

### Facet wrap

The facet_wrap() splits your chart based on a single variable. You define which variable to split upon with `~` followed by the variable name.

1. Add a new chunk and create the facet wrap shown here.

```{r facet-wrap}
five_plot +
  facet_wrap(~ state) +
  theme(legend.position = "none") # removes the legend. Try it without it!
```

A couple of notes about the above code:

- Note the comment in the code above where we used the `theme()` function to remove the legend.
- You can specify the number of rows or columns of the grouping by adjusting the facet_wrap() function: `facet_wrap(~ state, nrow = 2)` or `facet_wrap(~ state, ncol = 2)`. Try them!

### Facet grids

A `facet_grid()` allows you to plot on a combination of variables. We don't really have two numbers to compare in our higher education data so we'll show this with the `mpg` data we've used before.

1. Start a new section noting you'll try facet grid.
2. Add the chunk below and run it.

Explanations follow the chart.

```{r facet-grid}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

This chart is kinda hard to read, but let's try:

- Inside the mini charts, the best gas mileage is toward the top (from `hwy`) and the smaller engines are to the left (from `displ`.)
- The rows of charts are divided by drive train `drv`: four-wheel drive, front-wheel drive and rear-wheel drive.
- The columns of charts are divided by cylinders: like a 4-cylinder car vs 8-cylinder car.

This chart tells us that 4-cylinder, front-wheel drive cars with smaller engines get the best gas mileage. The blank charts mean that combination of values didn't exist in the data.

## On your own: Facet wrap

1. Create a section about doing a facet wrap on your own.
1. Take the "On your own" plot that you made earlier (The school spending for Gulf states) and apply a `facet_wrap()` here. You were instructed to save the plot into an R object, so you should be able to use that.
1. Remove the legend since each mini chart is labeled.


```{r homework-facet-example, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
ed_gulf_facet <- ed_gulf_plot +
  facet_wrap(~state) +
  theme(legend.position = "none")

ed_gulf_facet
```

## Saving plots

To save plots, you can right-click plots that you make in RNotebooks. Or, you can use the export button in the Plot pane. Or (and this is a preferred strategy), you can save them using `ggsave()`. ([Learn more here](https://ggplot2.tidyverse.org/reference/ggsave.html)).

```{r ggsave}
ggsave("images/txplot.png", plot = tx_plot)
```

Using `ggsave` creates a higher-res image than other methods.

## Interactive plots

Want to make your plot interactive? You can use [plotly](https://plotly.com/r/)'s `ggplotly()` function to transform your graph into an interactive chart.

To use plotly, you’ll want to install the plotly package, add the library, and then use the ggplotly() function:

1. In your R Console, run `install.packages("plotly")`. (You only have to do this once on your computer.)
1. Add a new section to note you are creating an interactive chart.
2. Add the code below and run it. Then play with the chart!

```{r plotly}
library(plotly)

tx_plot %>% 
  ggplotly()
```

Now you have tool tips on your points when you hover over them.

The `ggplotly()` function is not perfect. Alternatively, you can use plotly's own syntax to build some quite interesting charts, but it's a whole [new syntax to master](https://plotly.com/r/).

## What we learned

There is so much more to ggplot2 than what we've shown here, but these are the basics that should get you through the class. At the top of this chapter are a list of other resources to learn more.

