--- 
title: "Reporting with Data"
author: "Christian McDonald"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: utdata/rwd-class
description: "Reporting with Data is a class in the School of Journalism, Moody College of Communication at the University of Texas at Austin. It is taught by Christian McDonald."
---

# About this class

I'm a strong proponent of Scripted Journalism, a method of committing data-centric journalism in a programatic, repeatable and transparent way. There are a myriad of programming languages that further this, including R, Python (especially pandas and Jupyter) and [Observable](https://beta.observablehq.com/), but we'll be using R and RStudio.

## About the author

About me.

---

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
